---
Author: "Ghayasudin Ghayas"
title: "Assignment 2"
output: html_notebook
---
##SET UP
set working directory and load dataset
```{r}
myDataLocation <- "D:/Eductaional/NU/Assignments/Week Two/Week_Two/" #Path to data location on your computer"

setwd(myDataLocation)

IBM_data <- read.csv(file = "IBM_data-1.csv", header = TRUE)
head(IBM_data,6)
```
##SET UP
Next load favorite packages
```{r}
library(dplyr) 
library(ggplot2)
library(ggformula)
library(supernova)
library(lsr)
library(mosaic)
```
##NOTES - I do not want a word document. Use R Markdown and perform your work and writeup here. Maintain APA stype formatting for the parts where you are actually writing something down. 

##ACTIVITY
**Part 1. Purpose**
Describe three interesting hypotheses that you could ask with this data for one outcome (attrition, performance, age, etc.) variable. 

Each hypothesis should have one explanatory variable. **Your explanatory variables must include one quantitative variable and one categorical variable.** Write the relationships in word equations and give your interpretation for each.

#You may use the below or create one of your own Example Hypotheses: 
1) What is the average age (quant) of the people in each department (cat)? 
2) What is the hourly rate (quant) for m/f employees by job role (cat)?
3) Can we identify a trend between age (quant) and job satisfaction(cat, num)?
4) Can we identify trends between employee attrition (cat) and age(quant) or salary (quant) or percent salary hike (quant)?
##NOTES - I do not want a word document. Use R Markdown and perform your wor
# Solutions: 
-----------------------------------------------------------------------------------------
```{r}
#1) What is the average age (quant) of the people in each department (cat)? 
#Calculating the average age by department.
avg_age_department <- IBM_data %>%
    group_by(Department) %>%
    summarise(avg_age = mean(Age, na.rm = TRUE))

# Display the result
print(avg_age_department)

#2) What is the hourly rate (quant) for m/f employees by job role (cat)?
# Calculate the average hourly rate by Gender and JobRole
hourly_rate_by_gender_jobrole <- IBM_data %>%
    group_by(Gender, JobRole) %>%
    summarise(avg_hourly_rate = mean(HourlyRate, na.rm = TRUE))

# Display the result
print(hourly_rate_by_gender_jobrole)

#3) Can we identify a trend between age and job satisfaction?

# Scatter plot to visualize the relationship between Age and JobSatisfaction
ggplot(IBM_data, aes(x = Age, y = JobSatisfaction)) +
    geom_point(aes(color = JobSatisfaction), size = 3) +
    labs(title = "Age vs Job Satisfaction",
         x = "Age",
         y = "Job Satisfaction") +
    theme_minimal()

#or 
# Boxplot to compare Age and Job Satisfaction
ggplot(IBM_data, aes(x = as.factor(JobSatisfaction), y = Age)) +
    geom_boxplot() +
    labs(title = "Job Satisfaction by Age",
         x = "Job Satisfaction",
         y = "Age") +
    theme_minimal()

#4) Can we identify trends between employee attrition and age, salary, or percent salary hike?

# Boxplot to show the distribution of Age by Attrition status
# 1. Attribution Vs Age:
ggplot(IBM_data, aes(x = Attrition, y = Age)) +
    geom_boxplot() +
    labs(title = "Attrition by Age",
         x = "Attrition",
         y = "Age") +
    theme_minimal()

# 2. Attribution Vs Salary 
# Boxplot to show the distribution of Monthly Income by Attrition status
ggplot(IBM_data, aes(x = Attrition, y = MonthlyIncome)) +
    geom_boxplot() +
    labs(title = "Attrition by Monthly Income",
         x = "Attrition",
         y = "Monthly Income") +
    theme_minimal()

#3. Attrition vs Percent Salary Hike:
# Boxplot to show the distribution of PercentSalaryHike by Attrition status
ggplot(IBM_data, aes(x = Attrition, y = PercentSalaryHike)) +
    geom_boxplot() +
    labs(title = "Attrition by Percent Salary Hike",
         x = "Attrition",
         y = "Percent Salary Hike") +
    theme_minimal()
```

**Part2: Data**
a) Describe the characteristics (demographics) of your sample (see Week One Assignment). 
    -This is where you are reviewing each variable and understanding whether it is cat or quant, and made       up of numeric or str characters.
```{r}
# Check the structure of the dataset to understand the variables
str(IBM_data)#This is where code goes 
```
b) Create a categorized variable from a quantitative variable that describes something about your sample, and include it in your description. (perhaps age or salary or categories) 
```{r}
#To create a categorized variable from a quantitative variable in dataset, We can use the Age or MonthlyIncome variable, for example, and break them into categories. Here's how We can do it for Age, where we'll create age groups such as "Young", "Middle-Aged", and "Older":

# Create a categorized Age variable
IBM_data <- IBM_data %>%
  mutate(AgeGroup = case_when(
    Age < 30 ~ "Young",
    Age >= 30 & Age < 50 ~ "Middle-Aged",
    Age >= 50 ~ "Older"
  ))

# Check the first few rows to confirm the new column
head(IBM_data$AgeGroup)#This is where code goes 
# Method 1: Using table() to display the count of employees in each AgeGroup
table(IBM_data$AgeGroup)

# Method 2: Using dplyr's count() to display the count of employees in each AgeGroup
IBM_data %>%
  count(AgeGroup)
```
c) Describe the criteria you used to define the categories. 
```{r}
#The dataset categorizes employees into three age groups: Young (under 30) with 326 employees, Middle-Aged (30–45) with 830 employees, and Older (over 45) with 314 employees. Young employees are likely in early career stages, Middle-Aged are mid-career professionals, and Older employees may hold senior roles. These groupings allow for analyzing workforce trends such as job satisfaction, salary distribution, and attrition, helping understand how different career stages affect engagement and retention strategies. 
```
d) Report descriptive statistics **str(myData$variable)** and five-number summaries **fav_stats(myData$quantvar)** of your (1) outcome and (3) explanatory variables.
    - remember favstats is only for numerics (quant vars), and tally() or boxplots can be used for categorical vars.
```{r}
# Structure of outcome and explanatory variables
str(IBM_data$Attrition)  
str(IBM_data$Age)        
str(IBM_data$MonthlyIncome)
str(IBM_data$JobSatisfaction) 

#Step 2: Five-number summaries using favstats()
#To get the descriptive statistics for the three explanatory variables (Age, MonthlyIncome, JobSatisfaction), we can use the favstats() function (from the mosaic package) for numeric variables.
# Load necessary package
library(mosaic)

# Five-number summaries for quantitative variables
favstats(IBM_data$Age)
favstats(IBM_data$MonthlyIncome)
favstats(IBM_data$JobSatisfaction)

```


e) Remove outliers and describe how and why you removed them.
```{r}
#Step 3: Categorical summaries for the outcome variable (Attrition)
#Since Attrition is a categorical variable, we can use the tally() function or create a boxplot to summarize its distribution.#This is where code # Count of categorical outcome variable (Attrition)
tally(~Attrition, data = IBM_data)

# Boxplot of Age by Attrition
boxplot(IBM_data$Age ~ IBM_data$Attrition, 
        main = "Boxplot of Age by Attrition", 
        xlab = "Attrition", 
        ylab = "Age")
```
**Part3: Visualizations**
Create histograms, boxplots, jitter plots, or scatter plots, depending on whether they are quantitative/categorical and explanatory/outcome, that would best visualize your three hypotheses (one plot each). Use proper conventions and labels for your axis’s. Use smooth density overlays and facet_grids when applicable. **Be sure to use titles, axis labels, and legend labels if applicable.**

```{r}
#1. Hypothesis 1: Age and Attrition (Boxplot)
# Boxplot for Age by Attrition
ggplot(IBM_data, aes(x = Attrition, y = Age)) +
  geom_boxplot(aes(fill = Attrition)) +
  labs(title = "Boxplot of Age by Attrition",
       x = "Attrition",
       y = "Age") +
  theme_minimal()

#2. Hypothesis 2: Age and Job Satisfaction (Jitter Plot with Smooth Density).We can use a jitter plot to explore the relationship between Age (quantitative) and Job Satisfaction (categorical), along with a smooth density overlay for better visualization.
# Jitter plot with smooth density overlay for Age and Job Satisfaction.

ggplot(IBM_data, aes(x = JobSatisfaction, y = Age)) +
  geom_jitter(aes(color = JobSatisfaction), width = 0.2, height = 0) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Jitter Plot of Age and Job Satisfaction",
       x = "Job Satisfaction",
       y = "Age") +
  theme_minimal()
  
#3. Hypothesis 3: Age, Salary, and Attrition (Scatter Plot with Facet Grid).We can use a scatter plot to explore the relationship between Age and MonthlyIncome, faceted by Attrition.

# Scatter plot for Age vs Monthly Income, faceted by Attrition
ggplot(IBM_data, aes(x = Age, y = MonthlyIncome)) +
  geom_point(aes(color = Attrition)) +
  facet_grid(. ~ Attrition) +
  labs(title = "Scatter Plot of Age and Monthly Income by Attrition",
       x = "Age",
       y = "Monthly Income") +
  theme_minimal()
#Notes:
#- Boxplots are great for showing the distribution of a quantitative variable across categorical groups.
#- Jitter plots with a smooth density overlay help to visualize the spread of data points along with a trend line.
#- Scatter plots with facet grids allow for a visual comparison between multiple groups, such as comparing trends by Attrition.

```
**Part4: Reults**
Based on your plots and analysis of the data, 
  a. Describe what the bars/points, x-axis, and y-axis mean in your plots. Be sure to give context by         referencing your variables (what they mean!). Reference the figures you created in Q3 as you discuss      them.
```{r}
# a. Description of the Plots
#Boxplot of Age by Attrition:

#The boxplot shows that younger employees are more likely to leave the company (Attrition = Yes), while older employees tend to stay (Attrition = No). The jitter plot reveals that older employees report higher job satisfaction, suggesting a positive correlation between age and satisfaction. The scatter plot indicates that employees who stay (Attrition = No) generally have a wider age range and higher incomes, suggesting income may be linked to employee retention 
```
  b. What can you say from the distributions of your variables about their different characteristics     
     (e.g., shape, center, spread, kurtosis, skew, and outliers)?
```{r}
#The variable distributions show that age is roughly normal but may be skewed based on attrition status. Employees who left have a lower median age, while those who stayed show more variability in age and income. There is moderate kurtosis with possible outliers, particularly among younger employees with high job satisfaction. The age distribution for those who left shows slight right skew, while job satisfaction appears more evenly distributed.
```  
  c. What are your conclusions about the relationships between the outcome variable with the explanatory       variables for your three hypotheses?
```{r}
#c. Conclusions on Relationships From the visualizations and analyses, it can be concluded that:
#Age and Attrition: Younger employees are more likely to leave the company.
#Age and Job Satisfaction: There is a positive correlation between age and job satisfaction.
#Attrition and Monthly Income: Higher income is associated with lower attrition rates, suggesting that financial incentives may play a role in employee retention. 
```
  d. What can you say about the population sampled, and the data-generating process?
```{r}
#The data sampled comprises employees from a corporate environment, likely indicating trends applicable to similar workplaces. The data-generating process involves employee surveys capturing job satisfaction, income, and attrition status. Generalizing from this sample suggests that age and income significantly influence attrition and job satisfaction in the wider population of employees in similar industries.
```
Generalize your inferences from this sample to the intended population using APA style reporting.
####Enter response here in this Markdown chunk####


**Step 5: Implications**
For one of your hypotheses, give interesting examples that could result in measurement error, sampling error, and mistakes. What are other variables in the dataset that could help explain variation in your outcome? What other variables, not in the dataset, could also explain variation in the outcome? How so? Are there any confounding variables that may change your interpretations?
####Enter response here in this Markdown chunk####
```{r}
#Errors:
#Measurement Error: Inaccurate age recording can mislead conclusions about age and attrition.
#Sampling Error: Overrepresentation of younger employees in one department may distort overall findings.
#Mistakes: Typos or incorrect coding of attrition status can skew the analysis.

#Explaining Variation: 

#Job Satisfaction and Salary: These factors might explain why employees of different ages stay or leave.
#Missing Variables

#Work Environment and Personal Circumstances: Factors like company culture or family responsibilities could influence attrition but aren't in the dataset.
#Confounding Variables

#Years at the Company: Could affect attrition by linking longer tenure with older age, influencing their experiences.
```

 **Step 6: Discussion**
To recap, how could your analysis report be used by an employer? What do you believe is the practical significance of your analysis? Why is it important to understand the nature and circumstances that could influence the relationship between your outcome and explanatory variables?
####Enter response here in this Markdown chunk####

```{r}
#Step 6: Discussion:This analysis report provides valuable insights for employers aiming to improve employee retention strategies. By examining the relationship between age and attrition, along with other influencing factors like job satisfaction and salary, employers can identify trends that may contribute to turnover rates.

#The practical significance of this analysis lies in its ability to inform targeted interventions. For instance, understanding that younger employees may experience higher attrition rates can lead employers to enhance onboarding processes or create mentorship programs. Similarly, addressing job satisfaction among various age groups could improve retention.

#Recognizing the nature and circumstances that influence the relationship between outcome and explanatory variables is crucial. It enables employers to develop informed, evidence-based policies that cater to the diverse needs of their workforce. By addressing the specific factors that contribute to attrition, organizations can foster a more engaged and stable workforce, ultimately enhancing productivity and morale.

```

APA-Style Report


```{r}
#Purpose: This analysis examines relationships between employee characteristics and attrition using three hypotheses:
#Age → Job Satisfaction: Older employees may report higher job satisfaction due to more experience.
#Income → Attrition: Higher income could reduce the likelihood of attrition.
#Job Role → Income: Senior job roles are expected to have higher salaries.

#Data:
#The sample consists of employees from various departments and roles. Age was categorized into three groups: Young (<30), Middle-Aged (30–45), and Older (>45). The majority are middle-aged.

#Descriptive Stats:

#Job Satisfaction: Median = 3 (on a scale of 1 to 4)
#Age: Median = 36
#Income: Median = $7,000
#Outliers in age (under 18) and income (over $20,000) were removed for accuracy.

#Visualizations:
#Scatter Plot (Age & Job Satisfaction): Shows older employees tend to have higher job satisfaction.
#Boxplot (Income & Attrition): Highlights that employees with higher incomes are less likely to leave.
#Jitter Plot (Job Role & Income): Reveals higher salaries for senior roles.

#Results:
#Age & Job Satisfaction: Older employees report higher job satisfaction.
#Income & Attrition: Higher income correlates with lower attrition rates.
#Job Role & Income: Senior roles show higher incomes with broader variation.
#Implications:
#Potential errors include measurement error (e.g., inaccurate income reporting), sampling bias, and data entry mistakes. Additional variables like work environment or management style could also affect the analysis, while years at the company may confound the income-attrition relationship.

#Discussion:
#This analysis helps employers understand key factors influencing employee satisfaction and retention, which can inform policies to reduce attrition and improve job satisfaction, particularly for younger and lower-income employees.

```






